---
title: "Predicting Patient Length of Stay and Cost per Day"
author: "Courtney Gale & Randy Parrish"
date: "12/4/2020"
output: html_document
---

## Introduction

In this notebook, we aim to predict patient length of stay and cost per day. Patient length of stay is an important factor in hospital logistics. Knowing how long patients may need a room allows healthcare workers better allocate hospital resources. Length of stay also is an important metric that is used to evaluate a hospital's efficiency and the quality of care given to the patient. Predicting and understanding cost per day for a patient is also useful knowledge for a hospital and public health in general because understanding what siutations lead to high healthcare costs can inform future decisions on preventive care. Preventive care has been shown to be one of the best ways to decrease healthcare costs.

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(root.dir = "~/Users/Courtney/Documents/Emory_3rd_Semester/CS - Machine Learning/CS534-Final")
```

```{r, message = FALSE}
library(ggplot2)
library(caret)
library(future)
library(spaMM)
library(gt)
```


```{r, cache = TRUE}
# Import the data
path  <- '../data/Hospital_Inpatient_Discharges__SPARCS_De-Identified___2015.csv'
dat <- read.csv(path, 
                colClasses = c(rep('factor',10),'character',rep('factor',21),'numeric', rep('factor',2), 'character', 'character'))
```

### Display the data

```{r}
data_sub <- head(dat)

gt_first6 <-
  gt(data_sub) %>%
    tab_header(title = "First 6 Rows of Hospital Inpatient Data")

gt_first6
```

### Features

```{r}
# Look at makeup of dataset
str(dat)

# Remove dollar sign from Total Costs and Total Charges and make numeric
dat$Total.Charges <- as.numeric(substring(dat$Total.Charges, 2))
dat$Total.Costs <- as.numeric(substring(dat$Total.Costs, 2))

# make Length.of.Stay numeric
dat[which(dat$Length.of.Stay == '120 +'), 'Length.of.Stay'] <- '121'
dat$Length.of.Stay <- as.numeric(dat$Length.of.Stay)

# Create cost per day variable
dat$Cost.per.Day <- dat$Total.Costs / dat$Length.of.Stay
```

## Figures using all the data

Length of stay goes from 1 to 121 (previously 120+) but the majority of the data lies between 1 and 30. With the most patients having stayed 2 days. 

```{r}
ggplot(dat, aes(x = Length.of.Stay)) + geom_histogram(binwidth = 1, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  ggtitle("Distribution of Length of Stay") +
  xlab("Length of Stay") +
  ylab("Frequency") + 
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40), limits = c(0,40)) +
  scale_y_continuous(labels = scales::comma)
```

Mean length of stay increases as age group increases. The interquartile range of length of stay also increases as age group increases.

```{r, message = FALSE}
ggplot(dat, aes(x = Age.Group, y = Length.of.Stay, fill = Age.Group)) + 
  geom_boxplot() +
  theme(legend.position = "none") +
  ggtitle("Length of Stay by Age Group") +
  xlab("Age Group") +
  ylab("Length of Stay") +
  scale_y_continuous(breaks = c(1, 10, 20), limits = c(0,20))
```

The following codes are Major Diagnosic Codes (MDC) for that generally correspond to a major organ system. 

$~$

0 Pre-MDC or Ungroupable

1 Diseases and Disorders of the Nervous System

2 Diseases and Disorders of the Eye

3 Ear, Nose, Mouth, Throat, and Craniofacial Diseases and Disorders

4 Diseases and Disorders of the Respiratory System

5 Diseases and Disorders of the Circulatory System

6 Diseases and Disorders of the Digestive System

7 Diseases and Disorders of the Hepatobiliary System and Pancreas

8 Diseases and Disorders of the Musculoskeletal System and Connective Tissue

9 Diseases and Disorders of the Skin, Subcutaneous Tissue and Breast

10 Endocrine, Nutritional and Metabolic Diseases and Disorders

11 Diseases and Disorders of the Kidney and Urinary Tract

12 Diseases and Disorders of the Male Reproductive System

13 Diseases and Disorders of the Female Reproductive System

14 Pregnancy, Childbirth and the Puerperium

15 Newborns and Other Neonates with Conditions Originating in the Perinatal Period

16 Diseases and Disorders of Blood, Blood Forming Organs and Immunological Disorders

17 Lymphatic, Hematopoietic, Other Malignancies, Chemotherapy and Radiotherapy

18 Infectious and Parasitic Diseases, Systemic or Unspecified Sites

19 Mental Diseases and Disorders

20 Alcohol/Drug Use and Alcohol/Drug Induced Organic Mental Disorders

21 Poisonings, Toxic Effects, Other Injuries and Other Complications of Treatment

22 Burns

23 Rehabilitation, Aftercare, Other Factors Influencing Health Status and Other Health Service Contacts

24 Human Immunodeficiency Virus (HIV) Infections

25 Multiple Significant Trauma

$~$

The longest average length of stay among MDC groups was the 0 MDC group which corresponds to Pre-MDC or Ungroupable

```{r}
dat$APR.MDC.Code2 <- as.numeric(as.character(dat$APR.MDC.Code))

ggplot(aes(x = APR.MDC.Code2, y = Length.of.Stay), data = dat) + stat_summary(fun = "mean", geom = "bar",  fill = "#69b3a2") +
  ggtitle("Average Length of Stay by APR MDC Code") +
  xlab("APR MDC Code") +
  ylab("Average Length of Stay") +
  scale_x_continuous(breaks = seq(0, 25))
```

As severity of illness increases average length of stay increases. Cases labeled as extreme stay for an average of 15 days.

```{r}
dat$APR.Severity.of.Illness.Description <- as.character(dat$APR.Severity.of.Illness.Description)
dat$APR.Severity.of.Illness.Description <- factor(ifelse(dat$APR.Severity.of.Illness.Description == "", "NA", dat$APR.Severity.of.Illness.Description))

dat_sub <- dat[!dat$APR.Severity.of.Illness.Description == "NA", ]

ggplot(aes(x = reorder(APR.Severity.of.Illness.Description, Length.of.Stay), y = Length.of.Stay, fill = APR.Severity.of.Illness.Description), data = dat_sub) + stat_summary(fun = "mean", geom = "bar") +
  theme(legend.position = "none") +
  ggtitle("Length of Stay by Severity of Illness") +
  xlab("Severity of Illness") +
  ylab("Average Length of Stay")
```

Cost per day is highest in patients with elective procedures and cases of trauma.

```{r}
ggplot(aes(x = Type.of.Admission, y = Cost.per.Day, fill = Type.of.Admission), data = dat) + stat_summary(fun = "mean", geom = "bar") +
  theme(legend.position = "none") +
  ggtitle("Cost per Day by Type of Admission") +
  xlab("Admission Type") +
  ylab("Average Cost per Day")
```

Average cost per day increased as age group increased with the highest average cost in the age group from 50 to 69. Surprisingly, average cost for those 70 or older was less than both age groups of 30 to 49 and 50 to 69.

```{r}
ggplot(aes(x = Age.Group, y = Cost.per.Day, fill = Age.Group), data = dat) + stat_summary(fun = "mean", geom = "bar") +
  theme(legend.position = "none") +
  ggtitle("Cost per day by Age Group") +
  xlab("Age Group") +
  ylab("Average Cost per day")
```

Diseases and Disorders of the Male Reproductive System (MDC Code 12), Diseases and Disorders of the Musculoskeletal System and Connective Tissue (MDC Code 8), Diseases and Disorders of the Female Reproductive System (MDC Code 13) had the highest average costs per day. 


```{r}
ggplot(aes(x = APR.MDC.Code2, y = Cost.per.Day), data = dat) + stat_summary(fun = "mean", geom = "bar",  fill = "#69b3a2") +
  ggtitle("Average Cost per day by APR MDC Code") +
  xlab("APR MDC Code") +
  ylab("Average Cost per day") +
  scale_x_continuous(breaks = seq(0, 25))
```

## Filtering the Data

We chose to filter out observations that were related to pregnancy and newborn data as there are unique characteristics of their features. The birthweight feature, for example, was only reported for newborns. For all other observations, the birthweight was recorded as 0. With pregnancy related variables, some observations had a lot of missing data due to confidentiality issues in cases that may have been related to abortion (i.e. abortion procedure, miscarriage, ectopic pregnancy, etc.). Due to these unique characteritics of this data, we decided to filter out these observations.

Rows meeting the following criteria were filtered 

APR.MDC.Code = 14 ['Pregnancy, Childbirth And Puerperium']
APR.MDC.Code = 15 ['Newborn And Other Neonates (Perinatal Period)']
Abortion.Edit.Indicator = 'Y'
Type.of.Admission = 'Newborn'

```{r}

# rename Zip.Code...3.digits
colnames(dat)[which(colnames(dat) == 'Zip.Code...3.digits')] <- 'Zip'

######
# PREGNANCY-RELATED FILTERING

aprdrg <- unique(dat[,c('APR.DRG.Code','APR.DRG.Description')])
aprdrg <- aprdrg[order(aprdrg$APR.DRG.Code),]
rownames(aprdrg) <- NULL
# paste0(aprdrg[,1],'-',aprdrg[,2]) # view

ccs <- unique(dat[,c('CCS.Diagnosis.Code','CCS.Diagnosis.Description')])
ccs <- ccs[order(ccs$CCS.Diagnosis.Code),]
rownames(ccs) <- NULL
# paste0(ccs[,1],'-',ccs[,2]) # view

# terms for filtering out OB-related
obgyn_terms <- c('abortion', 'amniot', 'birth', 'C-section', 'delivery', 'fetal', 'fetopelvic', 'labor', 'liveborn', 'malposition', 'natal','neonat', 'obstetric', 'partum', 'pregnancy', 'OB-related', 'umbilical cord complication')
exclude_terms <- c('non-obstetric', 'except in labor')

# get APR.DRG.Code, CCS.Diagnosis.Code to filter out
aprdrg_obgyn_desc <- grep(paste0(obgyn_terms, collapse='|'), aprdrg$APR.DRG.Description, ignore.case=TRUE, value=TRUE)
aprdrg_obgyn_desc <- grep(paste0(exclude_terms, collapse='|') , aprdrg_obgyn_desc, ignore.case=TRUE, invert=TRUE, value=TRUE)
aprdrg_obgyn_codes <- aprdrg[aprdrg$APR.DRG.Description %in% aprdrg_obgyn_desc, 'APR.DRG.Code']

ccs_obgyn_desc <- grep(paste0(obgyn_terms, collapse='|'), ccs$CCS.Diagnosis.Description, ignore.case=TRUE, value=TRUE)
ccs_obgyn_desc <- grep(paste0(exclude_terms, collapse='|') , ccs_obgyn_desc, ignore.case=TRUE, invert=TRUE, value=TRUE)
ccs_obgyn_codes <- ccs[ccs$CCS.Diagnosis.Description %in% ccs_obgyn_desc, 'CCS.Diagnosis.Code']

# non-obgyn data
dat <- dplyr::filter(dat, !(CCS.Diagnosis.Code %in% ccs_obgyn_codes | APR.DRG.Code %in% aprdrg_obgyn_codes | APR.MDC.Code == "14" | APR.MDC.Code == "15" | Abortion.Edit.Indicator == 'Y' | Type.of.Admission == 'Newborn')  & Birth.Weight == 0)

dat <- dat[, which(!(colnames(dat) %in% c('Abortion.Edit.Indicator','Birth.Weight')))]

```

## Subset data 

Due to the amount of computing power that it would take to analyze the whole dataset, we chose to subset the data by the Primary Payment type of Medicare. 

```{r}
# Subset to only medicare patients
dat <- subset(dat, dat$Payment.Typology.1 == "Medicare")

# Split training and testing set 
set.seed(123)
intrain <-createDataPartition(y=dat$Cost.per.Day,p=0.7,list=FALSE)
training <-dat[intrain,]
testing <-dat[-intrain,]
```

## Figures using subset of data (Only Medicare)

The distribution of length of stay in the subset with only patients with Medicare resembles the overall distribution, but the largest frequency of length of stay in this subset is 3 days.

```{r}
ggplot(dat, aes(x = Length.of.Stay)) + geom_histogram(binwidth = 1, fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  ggtitle("Distribution of Length of Stay") +
  xlab("Length of Stay") +
  ylab("Frequency") + 
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40), limits = c(0,40)) +
  scale_y_continuous(labels = scales::comma)
```

The greatest number of medicare patients were admitted due to Diseases and Disorders of the Circulatory System (MDC Code 5) then  Diseases and Disorders of the Respiratory System (MDC Code 4) and Diseases and Disorders of the Musculoskeletal System and Connective Tissue (MDC Code 8).

```{r}
dat$APR.MDC.Code2 <- as.numeric(as.character(dat$APR.MDC.Code))

ggplot(aes(x = APR.MDC.Code2), data = dat) + geom_bar(fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  ggtitle("Frequency of APR MDC Code in Medicare Patients") +
  xlab("APR MDC Code") +
  ylab("Frequency") +
  scale_x_continuous(breaks = seq(0, 25))
```

As expected, the majority of patients with Medicare are 70 or older and also many in the 50 to 69 age group.

```{r}
ggplot(aes(x = Age.Group), data = dat) + geom_bar(fill="#69b3a2", color="#e9ecef", alpha=0.9) +
  ggtitle("Frequency of Individuals with Medicare by Age Group") +
  xlab("Medicare") +
  ylab("Frequency") +
  scale_y_continuous(labels = scales::comma)
```

## Analysis

Mean absolute error, mean relative error and root mean squared error were used as evaluation metrics for comparing our models.

```{r}
# Error functinos
MAE <- function(pred, actual) {mean(abs(pred - actual))}
MRE <- function(pred, actual) {mean(abs(pred - actual)/abs(actual))}
RMSE <- function(pred, actual) {sqrt(mean((actual - pred)^2))}
```


### Length of Stay

#### Linear Regression
We wanted to focus our efforts in predicting length of stay through regression methods and not classification methods because knowing the exact number of days a patient would need to stay would be overall more useful than knowing a range of days a patient may stay. We start with a simple linear regression as our base model although we acknowledge that fundamentally, this is not the proper model that we want to use to model length of stay. 

```{r}
linearReg <- lm(Length.of.Stay ~ Age.Group
                + Gender
                + Race
                + Ethnicity
                + Type.of.Admission
                + APR.MDC.Code
                + APR.Severity.of.Illness.Code
                + APR.Risk.of.Mortality
                + Payment.Typology.1
                + Emergency.Department.Indicator
                + Health.Service.Area
                + Hospital.County 
          #      + Facility.Id
              , data = training)
summary(linearReg)

linearReg_preds <- predict(linearReg, testing)

# Evaluate
linearReg_MAE <- MAE(linearReg_preds, testing$Length.of.Stay)

linearReg_MRE <- MRE(linearReg_preds, testing$Length.of.Stay)

linearReg_RMSE <- RMSE(linearReg_preds, testing$Length.of.Stay)

```


#### Poisson Model
When you look at thet possible values of length of stay, you see that length of stay can only be integer values greater than 1. Linear regression, on the other hand, will give you continuous values from -infinty to infinity. Therefore, our next step was to model length of stay as a Poisson regression which models count data. This still isn't quite the right model as a normal Poisson can take on values of 0. Since this is inpatient data, there will never be a patient who has a length of stay as 0.

```{r}
Poi <- glm(Length.of.Stay ~ Age.Group
                        + Gender
                        + Race
                        + Ethnicity
                        + Type.of.Admission
                        + APR.MDC.Code
                        + APR.Severity.of.Illness.Code
                        + APR.Risk.of.Mortality
                        + Payment.Typology.1
                        + Emergency.Department.Indicator
                        + Health.Service.Area
                        + Hospital.County
                   #     + Facility.Id , 
                        family = Poisson(link = 'log'),
                        data = training)
summary(Poi)

Poi_preds <- predict(Poi, testing)

# Evaluate
Poi_MAE <- MAE(Poi_preds, testing$Length.of.Stay)
Poi_MRE <- MRE(Poi_preds, testing$Length.of.Stay)
Poi_RMSE <- RMSE(Poi_preds, testing$Length.of.Stay)

```


#### Zero-Truncated Poisson Model
Next we look at the Zero-Truncated Poisson model. This is a modification of the Poisson model, and it can only take integer values from 1 to infinity. Therfore, this model is more appropriate to model length of stay because it will never take 0 values.

```{r}
# GLM with zero-truncated Poisson
model1 <- glm(Length.of.Stay ~ 
		+ Age.Group
		# + Zip
		+ Gender
		+ Race
		+ Ethnicity
		+ Type.of.Admission
		# + CCS.Diagnosis.Code
		# + CCS.Procedure.Code
		# + APR.DRG.Code
		+ APR.MDC.Code
		+ APR.Severity.of.Illness.Code
		+ APR.Risk.of.Mortality
		+ Payment.Typology.1
		# + Payment.Typology.2
		# + Payment.Typology.3
		+ Emergency.Department.Indicator
		+ Health.Service.Area
		+ Hospital.County 
		# + Facility.Id,
		,
	family = Tpoisson(link = 'log'),
	data = training)
summary(model1)



model1_2 <- glm(Length.of.Stay ~ 
		+ Age.Group
		# + Zip
		+ Gender
		+ Race
		+ Ethnicity
		+ Type.of.Admission
		# + CCS.Diagnosis.Code
		# + CCS.Procedure.Code
		# + APR.DRG.Code
		+ APR.MDC.Code
		+ APR.Severity.of.Illness.Code
		+ APR.Risk.of.Mortality
		+ Payment.Typology.1
		# + Payment.Typology.2
		# + Payment.Typology.3
		+ Emergency.Department.Indicator
		+ Health.Service.Area
		+ Hospital.County 
		+ Facility.Id,
		,
	family = Tpoisson(link = 'log'),
	data = training)
summary(model1_2)
```


#### Hierarchical Zero-Truncated Poisson Model
Lastly, we fit a hierarchical zero-truncted poisson model. If you look at the variables for Health Serivce Area, County, and Facility, you can see that these are nested varaibles. For example, specific facility ids will only occur in specific counties which only occur in specific health areas. By modeling length of stay as a multi-level model, we can take into account the area-level, county-level, and facility-level effects on length of stay by accounting for unexplained variation in hospital service areas, hospital counties, and facilities. 

```{r}
# GLM with zero-truncated Poisson, nesting
model2 <- HLfit(Length.of.Stay ~ 
		+ Age.Group
		# + Zip
		+ Gender
		+ Race
		+ Ethnicity
		+ Type.of.Admission
		# + CCS.Diagnosis.Code
		# + CCS.Procedure.Code
		# + APR.DRG.Code
		+ APR.MDC.Code
		+ APR.Severity.of.Illness.Code
		+ APR.Risk.of.Mortality
		+ Payment.Typology.1
		# + Payment.Typology.2
		# + Payment.Typology.3
		+ Emergency.Department.Indicator
		+ Health.Service.Area
		+ (1 | Hospital.County / Health.Service.Area)
		,
	family = Tpoisson(link = 'log'), 
	data = training)
summary(model2, details=TRUE, max.print=99999999)


# GLM with zero-truncated Poisson, nesting
model3 <- HLfit(Length.of.Stay ~ 
		+ Age.Group
		# + Zip
		+ Gender
		+ Race
		+ Ethnicity
		+ Type.of.Admission
		# + CCS.Diagnosis.Code
		# + CCS.Procedure.Code
		# + APR.DRG.Code
		+ APR.MDC.Code
		+ APR.Severity.of.Illness.Code
		+ APR.Risk.of.Mortality
		+ Payment.Typology.1
		# + Payment.Typology.2
		# + Payment.Typology.3
		+ Emergency.Department.Indicator
		+ Health.Service.Area
		+ (1 | Hospital.County / Health.Service.Area)
		+ (1 | Facility.Id / Hospital.County / Health.Service.Area)
		,
	family = Tpoisson(link = 'log'), 
	data = training)
summary(model3, details=TRUE, max.print=99999999)
```


### Cost per Day

### Linear Regression
In predicting cost per day, we focused on comparing simple linear regression results to the results for a hierarchical linear model accounting for the variance in serivce area, county, and facility. We hoped and predicted that we would see an improvement in performance by using a more complicated and computationally heavy approach using the hierarchical structure. 

```{r}
# Linear regression for cost per day
linearReg <- lm(Cost.per.Day ~ Age.Group
                + Gender
                + Race
                + Ethnicity
                + Type.of.Admission
                + APR.MDC.Code
                + APR.Severity.of.Illness.Code
                + APR.Risk.of.Mortality
                + Emergency.Department.Indicator
                + Health.Service.Area
                + Hospital.County 
                + Facility.Id
                , data = training)
summary(linearReg)
```

63 coefficients were not estimated due to singularities. These all occured with hospital counties and facility ids. This is likely due to the nested strucutre of the data because each facility id only occurs within a single hospital county which only occurs within a single health service area. 

The output shows a R-squared of 0.2547 on the training data. 

```{r}
#Evaluation Metrics
#testing <- testing[!testing$Health.Service.Area == "",]

# Remove observations when cost per day are 0 because mean relative error will give infinity for those observations
testing <- testing[!testing$Cost.per.Day == 0, ]

linearReg_preds <- predict(linearReg, testing)

linearReg_MAE <- MAE(linearReg_preds, testing$Cost.per.Day)
linearReg_MRE <- MRE(linearReg_preds, testing$Cost.per.Day)
linearReg_RMSE <- RMSE(linearReg_preds, testing$Cost.per.Day)
```



#### Hierarchical Linear Regression

```{r}
Hierarchical_cost <- HLfit(Cost.per.Day ~  Age.Group
             + Gender
             + Race
             + Ethnicity
             + Type.of.Admission
             + APR.MDC.Code
             + APR.Severity.of.Illness.Code
             + APR.Risk.of.Mortality
        #     + Payment.Typology.1
        #     + Payment.Typology.2 
        #     + Payment.Typology.3
             + Emergency.Department.Indicator
             + Health.Service.Area
             + (1 | Hospital.County / Health.Service.Area)
             + (1 | Facility.Id / Hospital.County / Health.Service.Area), 
             family = gaussian, data = training)

summary(Hierarchical_cost)
```

```{r}
# Evaluation metrics
hierarchical_preds <- predict(Hierarchical_cost, testing)

hierarchical_MAE <- MAE(hierarchical_preds, testing$Cost.per.Day)
hierarchical_MRE <- MRE(hierarchical_preds, testing$Cost.per.Day)
hierarchical_RMSE <- RMSE(hierarchical_preds, testing$Cost.per.Day)
```

# Compare Models
```{r}
compare <- data.frame(matrix(, nrow = 4, ncol = 3))
colnames(compare) <- c("", "Linear Regression", "Hierarchical Linear Regression")

#Factor
compare[1,1] <- "Mean Absolute Error"
compare[2,1] <- "Mean Relative Error"
compare[3,1] <- "Root Mean Squared Error"

#Linear Regression
compare[1,2] <- linearReg_MAE
compare[2,2] <- linearReg_MRE
compare[3,2] <- linearReg_RMSE
  
# Hierarchical Linear Regression
compare[1,3] <- hierarchical_MAE
compare[2,3] <- hierarchical_MRE
compare[3,3] <- hierarchical_RMSE

# Print table
compare %>%
  kbl() %>%
  kable_styling()
```

